Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-20 paper/paper.tex
still on liveness

\documentclass{sig-alternate}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}

@@

%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% 1st. author
\alignauthor K. Prasanna Kumar \\
       \affaddr{IIT Bombay,}\\
       \affaddr{Mumbai 400076, India}
       \email{prasanna@cse.iitb.ac.in}
% 2nd. author
\alignauthor Amey Karkare\\
       \affaddr{IIT Kanpur}\\
       \affaddr{Kanpur 208016, India}
       \email{karkare@cse.iitk.ac.in}
% 3rd. author
\alignauthor Amitabha Sanyal \\
       \affaddr{IIT Bombay}\\
       \affaddr{Mumbai 400076, India}\\
       \email{as@cse.iitb.ac.in}
}
\maketitle
\begin{abstract}
We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
functional  programming  languages. The  first  is  to modify  garbage
collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
reachable data}.  This results in an increase in the garbage reclaimed
and  consequently in  fewer  garbage collections.   The  second is  to
eagerly evaluate closures  that are guaranteed to be  evaluated in the
future.

@@

@@

A variable is {\em live} it  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
the  liveness of pointers  including fields  of $\CONS$  cells.  Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and third  members of $x$. In this  notation, a scalar variable
being live or dead are represented as $\{\epsilon\}$ and $\{\}$ respectively.

A {\em liveness environment} is a mapping from variables to subsets of
$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
before executing $e$.
%%
%% Due to the inherent uncertainty in
%% determining the  point of evaluation  in lazy languages,  we associate
%% liveness environments with certain kinds  of paths in the program. All
%% programs points lying on the path share the same liveness environment.
%% We explain this point later.
%% 
A  notion  related  to  liveness  is  {\em  demand}.   The  demand  on
expression  $e$   is  again  an   access  path---that  subset   of  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.   The  demand on  an  expression  generates  liveness for  its
constituent  variables.  So  for  example, given  a (non-null)  demand
$\sigma =  \lbrace \epsilon, \acar  \rbrace$ and the  expression $\pi:
(\CDR\  x)$, the liveness  at $\pi$  is $\lbrace  x.\epsilon, x.\acdr,
x.\acdr\acar \rbrace$.  This indicates  that to explore $\sigma$ parts
of the  result of $(\CDR\  x)$, one has  to first dereference  $x$ and
\CDR\  field of  the cons  cell referred  by $x$.  Note  that liveness
refers to  variables and  fields, and not  to $\CONS$ cells  (i.e.\ to
edges  in  the memory  graph,  not  to  locations themselves).   Hence
liveness of $\{x.\epsilon,x.\acar\}$ means that future computation may
refer to  the value  $\ell$ of  variable $x$, and  also to  the $\CAR$
field of location $\ell$.

%% The classical  analogy of this  is in
%% {\em strong liveness},  which can be expressed in  our terminology as:
%% $y$ and $z$  have a liveness of $\lbrace\epsilon\rbrace$  at the entry
%% to $n: x:=y+z$,  c$y$ and $z$ if  (and only if) $x$ has  a liveness of
%% $\lbrace\epsilon\rbrace$ at exit of  $n$---the liveness of $x$ at exit
%% from $n$ becomes the demand on $y+z$.

%% Note  that, for  an  operation  like division  which  may raise  an
%% exception,  the  assignment $n:  x:=y/z$  makes  $y$  and $z$  live
%% regardless of the liveness of $x$.

We use $\sigma$  to range over demands, $\alpha$  to range over access
paths  and $\Lv$  to range  over liveness  environments.  The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in  \sigma_1, \alpha_2 \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$   is   a   shorthand  for   $\lbrace\acar\rbrace\sigma$.
Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
These transform demands on a function call into demands for its formal
parameters:    given    a    user    function    $f$,    defined    by
$(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
$\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.

%% Note that liveness refers to  variables and fields, and not to $\CONS$
%% cells  (i.e.\  to  edges  in   the  memory  graph,  not  to  locations
%% themselves).   Hence liveness  of $\{x.\epsilon,x.\acar\}$  means that
%% future computation may refer to  the value $\ell$ of variable $x$, and
%% also to the $\CAR$ field of  location $\ell$.  In the absence of other
%% pointers to heap location $\ell$, we are certain that the $\CDR$ field
%% of  $\ell$  will  not  be   referenced  and  may  hence  be  corrupted
%% arbitrarily.   Note therefore,  that  while $\ell$  cannot be  garbage
%% collected, any location  $\ell'$ stored in the $\CDR$  field of $\ell$
%% would  be  garbage (again  provided  there  are  no other  aliases  to
%% $\ell$).

\bigskip
\hrule 
\bigskip

\subsection{Liveness Analysis for lazy languages}
\comment{analysis for liveness has two components - transfering
  demands on expressions to liveness of results
and propagating liveness. first discuss demand --> liveness}

%%Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
%%analogous.
% has three analogous aspects.
% \begin{itemize}
% \item
\alan{Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
returns the  incremental liveness {\em generated} by  $s$.  Because we
generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
determine which access paths of its free variables are made live.  The
cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
% The cases for
% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
A  demand of  $\sigma$  on  (\CAR~$x$) is  transformed  to the  demand
$\acar\sigma$  on $x$.   In  addition, \CAR\  always dereferences  its
argument  (even if  its result  is  never used).   This generates  the
liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$ (note  $\sigma$  may be
$\{\}$).   In  the opposite  sense,  the  demand  of $\acar\sigma$  on
(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
that \CONS\ does not,  by itself, dereference its arguments.  Thirdly,
for the case  of a user-function call, a  third parameter $\Lfonly$ to
$\mathit{ref}$  expresses how  the demand  $\sigma$ on  the  result is
transformed  into demands  on its  parameters.  Constants  generate no
liveness.}

Liveness analysis for lazy languages differ from eager languages in
two respects:
\begin{enumerate}
\item Unlike eager languages, a variable may be bound to an expression
  that is not fully evaluated. i.e. a closure.
\item The point where an expression gets evaluated cannot be statically
  determined.
\end{enumerate}

To  deal  with the  first  issue, we  apply  the  results of  liveness
analysis only  to fully evaluated  data.  If the marking  phase during
garbage collection leads to a closure, and the closure is not dead, we
mark the entire reachable memory  under the closure for collection.  A
better scheme  would be  to garbage collect  the closure based  on the
liveness  at the point  of its  creation.  This  is sound  because our
language is statically scoped.  However, this is work in progress, and
the method described in this paper switches over to reachability while
collecting closures.

%% However, we GC,  if we find a variable can either  point to a fully
%% evaluated  data value  or  it could  point  to a  closure which  on
%% evaluation  generates  the   required  data  value.   For  example,
%% consider the statement {\tt  let \py~ $\leftarrow $ (\CDR~\px)}. In
%% case of  an eager  language the variable  \py~ points to  the \CDR~
%% part of  \px~ which  is guaranteed  to be a  cons cell.   Hence any
%% demand on \py~ can be  immediately transferred to the \CDR~ part of
%% \px.  In case  of a lazy language, \px~ might  not be evaluated and
%% hence  pointing to  a closure.  In our  example, it  points  to the
%% closure corresponding to (\CONS~1~\px).  It does not make any sense
%% to transfer the demand onto the \CDR~ part of \px~ as it still does
%% not exist.  , 2) it is  not possible to statically determine when a
%% closure gets evaluated.  Therefore during  GC we treat a closure as
%% a demand transformer and tranform the demand on to the arguments of
%% the  closure   pointed.   This  is   similar  to  the   process  of
%% transferring the demand from the  return value of a function to the
%% arguments of the function call in an eager language.

%% In our example, during GC if  we find that \px~ points to a closure
%% instead  of a  cons cell  we find  out the  demand/liveness  on the
%% arguments  of the  closure, namely  \py.  Statically  it  cannot be
%% determined  what  values  a  closure  points to  or  when  it  gets
%% evaluated, hence  our method  approximates the liveness  of closure
%% arguments  at  runtime  by  tagging  extra  information  with  each
%% closure.  Since  the arguments of the closures  could themselves be
%% closures in turn, we recursively  apply the same procedure until we
%% hit  a data  value.  The only  case  where we  do  not follow  this
%% procedure is if the demand on a particular closure is null. In such
%% a case we safely collect the entire closure as garbage.

To see  our way of  handling the second  issue, consider the liveness
analysis of the following function:

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} (\DEFINE\ (\pf~\pa)
	  \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
	  \UNL{3}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
          \UNL{4} \hspace*{.05cm}     $\pi_3\!\!:\,
          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
	  \UNL{5}  \hspace*{.05cm}    $\pi_4\!\!:\,
          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
	  \UNL{6} \hspace*{1.15cm}    $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
	  \UNL{7}  \hspace*{1.15cm}   $\pi_8\!\!:\,(\RETURN_1~\pu)$))
          \UNL{5} \hspace*{.05cm} $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
	\end{uprogram}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\medskip

@@

  The processing  {\LET} rule has an elegant formulation. Recollect
  that our language permits lazy \LET s whose definitions can be
  recursive   Thus, in the expression
  $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in $s$.  We model a let statement as a recursive function which captures an
  arbitrary  number
  of unrollings. We pass to this function the demand arising out of
  the liveness of $x$  at the
  beginning of $e$, and let the $the  
  that is the  most complex among  all the
  rules. A  {\LET} does  not add  any new label  but might  modify the
  liveness values of  some variables or might add  liveness values for
  the let bound variable. Since there might be more than one {\RETURN}
  statement following a {\LET},  there might be multiple {\em E-paths}
  reaching  the {\LET}.   The  demand  on the  let  variable might  be
  different along  different {\em E-paths}. Therefore,  for each label
  {\em  l}  in  $\cal{M}$,  the  liveness  of  the  let-expression  is
  calculated  using the  liveness environment  $\Lv$  corresponding to
  {\em l}.  Since the  {\LET} is lazy,  calculating its  livness might
  require a fixed-point processing.

\bigskip
\hrule 
\bigskip

@@

The  function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
but note the liveness $x.\epsilon$  generated by the latter.  The case
$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
instruction:  the liveness  of $e$  is given  by taking  the liveness,
$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
$\mathit{ref}(s,\Lv(z),\Lfonly)$.

% \item
%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}
Finally, the  judgement form $\mathit{Prog}  \len \Lfonly$ is  used to
determine  $\Lfonly$\@.   Analogously   to  classical  liveness  being
computed  as  a  solution  of  dataflow  equations,  we  require,  via
inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
fixed-point property that:  when we assume $\Lfonly$ to  be the family
of demand transformers for the program then the calculated liveness of
each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
$\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
prefer  the  least  one  as  giving  the  least  liveness  subject  to
safety---and hence greatest amount of garbage collected.
% \end{itemize}

@@

%\comment{

%}

% \item

@@

By the time the program starts executing, the liveness automata are created and stored in a file. 
This is re-used every time the program executes. 

@@

An important difference in GC of lazy language compared to an eager language is the point of 
garbage collection. In case of eager languages only \CONS~ points and function calls are 
the GC points. In a lazy language every \LET~ statement is a potential GC point. Hence we need to store 
liveness automata corresponding to every \LET~ statement in the program. This leads to an explosion 
in the number of automata. From our observation that along an {\em E-path} all program points have 
the same liveness, we can reduce the number of automata state required by sharing the same automata 
for all program points along an {\em E-path}. In the example program, we only need to create automata 
for program points \ldots, since they are the only distinct {\em E-path}s in the program. 

@@

During GC, if the heap cell being processed is a cons cell, we consult the liveness automaton 
corresponding to it and copy only the live paths. In case the heap cell points to a closure, 
we safely approximate the liveness to be {\{0+1\}\*} and copy everything under the closure.

@@

A consequence of copying everything under a closure is the fact that we can end up trying to copy 
things which are not live and corrupt the heap. This can happen if at some $GC_k$ a cons cell is copied 
using liveness. At a later stage this cons cell becomes part of a closure and hence during $GC_{k+n}$ the 
algorithm tries to copy everything under the cons cell thus copying heap locations which might not be valid. 
To avoid this behavior we either have to choose to do a full liveness based GC for closure or we should 
ensure that any non-live references are not carried across GC's. We take the latter approach and ensure that 
non-live references are set to null. After every GC, the live part of the buffer is scanned and any reference 
which is not live is set to null. The correctness of this follows from the fact that any reference which is 

>>>>

\documentclass[10pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
\cmt{{
\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% 1st. author
\alignauthor K. Prasanna Kumar \\
       \affaddr{IIT Bombay,}\\
       \affaddr{Mumbai 400076, India}
       \email{prasanna@cse.iitb.ac.in}
% 2nd. author
\alignauthor Amey Karkare\\
       \affaddr{IIT Kanpur}\\
       \affaddr{Kanpur 208016, India}
       \email{karkare@cse.iitk.ac.in}
% 3rd. author
\alignauthor Amitabha Sanyal \\
       \affaddr{IIT Bombay}\\
       \affaddr{Mumbai 400076, India}\\
       \email{as@cse.iitb.ac.in}
}
}}
\authorinfo{Double Blind Review}{}{}
\maketitle

\begin{abstract}
We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
functional  programming  languages. The  first  is  to modify  garbage
collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
reachable data}.  This results in an increase in the garbage reclaimed
and  consequently in  fewer  garbage collections.   The  second is  to
eagerly evaluate closures  that are guaranteed to be  evaluated in the
future.

@@

\pagebreak

@@

A variable is {\em live} it  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
the  liveness of pointers  including fields  of $\CONS$  cells.  Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second  and  third  members  of  $x$. In  this  notation,  $x  \mapsto
\{\epsilon\}$ represents access using  $x$ itself and $x \mapsto \{\}$
indicates that $x$ is dead.  A {\em liveness environment} is a mapping
from  variables  to   subsets  of  $\{\acar,\acdr\}^\ast$,  but  often
expressed as  a set, for example by  writing $\{x.\acar\acdr, x.\acdr,
y.\epsilon\}$               instead               of               $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  
program  points $\pi$.  The liveness  at $\pi:e$  is the  liveness just
before executing $e$.

A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
expression  $e$  is again  a  set  of  access paths---that  subset  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\RETURN~x)$. Assume that the
context of  this expression produces the liveness  $[x \mapsto \lbrace
  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
  classical analogue  is strong liveness  analysis, where $y$  and $z$
  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
  the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).

We use $\sigma$  to range over demands, $\alpha$  to range over access
paths  and $\Lv$ to  range over  liveness environments.   The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
%% Finally,  we use $\Lfonly$  to range  over {\em  demand transformers}.
%% These transform demands on a function call into demands for its formal
%% parameters:    given    a    user    function    $f$,    defined    by
%% $(\DEFINE\  (f\ x_1\  \ldots\ x_n)\  \  e_f)$ and  called with  demand
%% $\sigma$, then $\Lf{f}{i}{\sigma}$ is the liveness of $x_i$ at $e_f$.

\subsection{Liveness Analysis for lazy languages}

Figure~\ref{fig:live-judge}  describes  our  analysis.   Firstly,  the
function $\mathit{ref}$,  when given an  application $s$ and  a demand
$\sigma$,  returns the liveness  generated for  the free  variables of
$s$.   In  each  case,  the  null demand  ($\emptyset$)  generates  no
liveness.   Notice  that this  is  different  from  liveness of  eager
languages, where, for instance,  (\CDR~$x$) would be evaluated even if
the   demand   on   it   is  $\emptyset$   generating   the   liveness
$\{x.\epsilon\}$.  The  cases (\PRIM~x~y)  and (\NULLQ~x) are  easy. A
non-null  demand of  $\sigma$  on (\CDR~$x$),  is  transformed to  the
liveness  $x.\epsilon,  x.\acdr\sigma$.   In  an opposite  sense,  the
demand  of  $\acar\sigma$ on  (\CONS~$x$~$y$)  is  transformed to  the
demand  $\sigma$  on $x$.   Note  that  \CONS\  does not,  by  itself,
dereference its arguments. Constants generate no liveness.

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program. $\Lfonly$ expresses how the  demand $\sigma$ on the result of
a function call is transformed  into liveness on its parameters at the
beginning of the call.  $\Lfonly$  is determined by the judgement form
$\mathit{Prog}    \len   \Lfonly$    using   inference    rule   ({\sc
  live-define}). This states the  fixpoint property to be satisfied by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated from their bodies. The algorithmic formulation of this rule
through  a  grammar will  be  described  in [???].   This yields  the  least
solution  satisfying  the  fixpoint  property  and  thus  ensures  the
greatest amount of garbage collected.

%%   Analogously   to  classical  liveness  being
%% computed  as  a  solution  of  dataflow  equations,  we  require,  via
%% inference   rule  ({\sc  live-define}),   $\Lfonly$  to   satisfy  the
%% fixed-point property that:  when we assume $\Lfonly$ to  be the family
%% of demand transformers for the program then the calculated liveness of
%% each function  body $\mathcal{L}(e_f,\sigma,\Lfonly)$ agrees  with the
%% assumed $\Lfonly$.   As usual, there  are often multiple  solutions to
%% $\Lfonly$;  all are  safe (see  Section~\ref{sec:correctness})  but we
%% prefer  the  least  one  as  giving  the  least  liveness  subject  to
%% safety---and hence greatest amount of garbage collected.

We now describe the  propagation of liveness through expressions. This
brings  us to  the  second way  in  which liveness  analysis for  lazy
languages  differ from  eager languages---the point where  an
expression gets evaluated cannot be statically determined.  To see how
this  is handled,  consider  the liveness  analysis  of the  following
function:
%%%%%%%%%%%%%%%%%%%%%%

% Push this to a later point

%% Unlike eager languages, a variable may be bound to an expression
%%   that is not fully evaluated. i.e. a closure.

%% To  deal  with the  first  issue, we  apply  the  results of  liveness
%% analysis only  to fully evaluated  data.  If the marking  phase during
%% garbage collection leads to a closure, and the closure is not dead, we
%% mark the entire reachable memory  under the closure for collection.  A
%% better scheme  would be  to garbage collect  the closure based  on the
%% liveness  at the point  of its  creation.  This  is sound  because our
%% language is statically scoped.  However, this is work in progress, and
%% the method described in this paper switches over to reachability while
%% collecting closures.

%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
          \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
          \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
	\end{uprogram}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\medskip

@@

  The  processing {\LET}  rule has  an elegant  formulation. Recollect
  that  our language  permits lazy  \LET  s whose  definitions can  be
  recursive Thus,  in the expression  $\LET~x \leftarrow~s~\IN~e$, $x$
  can occur  in $s$.   We model the  definition $x \leftarrow~s$  as a
  possibly recursive  function which  captures an arbitrary  number of
  unrollings of the  definition.  We pass to this  function the demand
  arising  out of the  liveness of  $x$ along  all paths  reaching the
  beginning of $e$, and let  the \Lfonly abstraction derive the demand
  on the  variables of $s$.

%% that is the  most complex among  all the
%%   rules.  A  {\LET} does not  add any new  label but might  modify the
%%   liveness values of  some variables or might add  liveness values for
%%   the  let  bound  variable.   Since  there might  be  more  than  one
%%   {\RETURN} statement following a {\LET}, there might be multiple {\em
%%     E-paths}  reaching the  {\LET}.  The  demand on  the  let variable
%%   might be different along different {\em E-paths}.

@@

\alan{The function  $\mathcal{L}$ now  gives  the  (total)  liveness of  an
expression $e$.   The cases $\RETURN$ and  $\SIF$ are straightforward,
but note the liveness $x.\epsilon$  generated by the latter.  The case
$(\LET\   z\leftarrow   s\  \IN\   e')$   resembles  a   three-address
instruction:  the liveness  of $e$  is given  by taking  the liveness,
$\Lv$, of $e'$, killing any liveness of $z$ and adding any incremental
liveness from  $s$.  The main subtlety  is how the liveness  of $z$ in
$\Lv$  is converted  to a  demand  $\Lv(z)$ to  be placed  on $s$  via
$\mathit{ref}(s,\Lv(z),\Lfonly)$.}

%%%%%%\marginpar{Say $\mathcal{L}$ is monotonic?}

% \end{itemize}

@@

% \item
\pagebreak

@@

By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.

@@

An important  difference in GC of  lazy language compared  to an eager
language  is  the  point  of  garbage collection.  In  case  of  eager
languages only \CONS~ points and  function calls are the GC points. In
a lazy language  every \LET~ statement is a  potential GC point. Hence
we  need  to store  liveness  automata  corresponding  to every  \LET~
statement in the program. This leads  to an explosion in the number of
automata. From our observation that  along an {\em E-path} all program
points have  the same liveness, we  can reduce the  number of automata
state required  by sharing  the same automata  for all  program points
along an {\em E-path}. In the  example program, we only need to create
automata for program  points \ldots, since they are  the only distinct
{\em E-path}s in the program.

@@

During GC, if the heap cell being processed is a cons cell, we consult
the  liveness automaton  corresponding to  it and  copy only  the live
paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
approximate the  liveness to be {\{0+1\}\*} and  copy everything under
the closure.

@@

A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closure or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
